---
title: "Using the DfE Colour Palette"
output:
  bookdown::html_document2:
    toc: true
    fig_caption: yes
pkgdown:
  as_is: true
  set_null_theme: false
params:
    n: NA
vignette: >
  %\VignetteIndexEntry{colourpalette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

The underlying DfE colour palette can be accessed at `usefulR::colour_palette`. It contains 6 colours (blue, red, orange, yellow, green and purple) in 5 different hues. The function `dfe_colours()` provides a simple way to create various colour schemes from the palette.

### Monochrome sequential schemes

A scheme which contains one colour in 5 different hues can be created by just using the `colour` argument.

```{r box}

# Box plot
bp <- ggplot(iris, aes(Species, Sepal.Length)) + 
  geom_boxplot(aes(fill = Species)) +
  theme_minimal()
bp + ggplot2::scale_fill_manual(values = usefulR::dfe_colours("blue"))

```

### Spectral schemes

```{r scatter}

# Scatter plot
sp <- ggplot(iris, aes(Sepal.Length, Sepal.Width)) + 
  geom_point(aes(color = Species)) +
  theme_minimal()
sp + ggplot2::scale_colour_manual(values = usefulR::dfe_colours(c("blue", "orange", "green"), intensity = 80))

```
