---
title: "Distributional Weights"
output:
  bookdown::html_document2:
    toc: true
    fig_caption: yes
pkgdown:
  as_is: true
  set_null_theme: false
params:
    n: NA
bibliography: refs.bib
link-citations: yes
vignette: >
  %\VignetteIndexEntry{Companion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
#library(usefulR)
devtools::load_all()
library(ggplot2)
library(kableExtra)
```

The @hm_treasury_green_book_green_2018 encourages the use of 'welfare' or 'distributional' weights. This is based on the concept of the diminishing marginal utility of income, where the value of the an additional pound is worth more to those with lower incomes than those on higher incomes. This short vignette discusses the theory, recent estimation, suggested best practice and provides a short example.

# The elasticity of marginal utility

If we assume a standard [isoelastic utility function](https://en.wikipedia.org/wiki/Isoelastic_utility):

$$
\begin{equation}
  u(y) = \begin{cases}
         \frac{y^{1-\rho}-1}{1-\rho} & \rho \neq 1 \\
         ln(y) & \rho = 1
         \end{cases}
  (\#eq:a)
\end{equation}
$$
Then the marginal utility of income is:

$$
\begin{equation}
  u'(y) = y^{-\rho}
  (\#eq:b)
\end{equation}
$$

And so the ratio between two marginal utilities is:

$$
\begin{equation}
  \frac{u'(y_1)}{u'(y_2)} = \left(\frac{y_2}{y_1}\right)^\rho
  (\#eq:c)
\end{equation}
$$

Therefore if $\rho = 1$ then marginal utility would be simply inversely proportional to earnings ($y$) implying an individual, for example earning £25,000 would gain twice as much utility from an extra pound than someone earning £50,000. However, most evidence suggests that $\rho$, the elasticity of marginal utility, is greater than 1. @layard_marginal_2008 estimate $\rho$ to be 1.30 (0.97-1.62) in the UK using the British Household Panel Survey (BHPS) and 1.26 (1.16-1.37) in a combined estimate using BHPS and 5 other international surveys. This implies marginal utility falls at an even faster rate with respect to income, as shown in figure \@ref(fig:plot). 

```{r plot, fig.cap = "Marginal Utility", fig.width = 6, fig.align = "center",  echo = FALSE, warning = FALSE}

eq1 = function(x){x^-1}
eq2 = function(x){x^-1.3}
ggplot(data.frame(x = c(1, 20)), aes(x = x)) + 
  stat_function(fun=eq1, size = 2, aes(colour = "1")) +
  stat_function(fun=eq2, size = 2, aes(colour = "2")) +
  scale_color_discrete(labels = c(expression(rho~" = 1"), expression(rho~"= 1.3"))) +
  xlab("Income") +
  ylab("Marginal Utility") +
  labs(colour = "") +
  theme_minimal() +
  theme(legend.text.align = 0)


```

The use of the @layard_marginal_2008 central estimate of $\rho = 1.3$ is suggested by @fujiwara_department_2010 and endorsed by the @hm_treasury_green_book_green_2018. 

# Welfare Weights

The @hm_treasury_green_book_green_2018's suggested methodology for calculating welfare weights is to split the household income distribution into **5 quintiles** and use the median of each quintile in calculations. The weights are derived as in equation \@ref(eq:c), by expressing the marginal utility of each quintile as a percentage of the marginal utility of the overall median.


# Example

We generate a hypothetical log-normal distribution of earnings:

```{r}
set.seed(1234)
earns <- rlnorm(10000) * 30000
summary(earns)
```

The 'welfare weights' can be estimated for a given $\rho$ using the `dist_weights` function:

```{r, echo = TRUE}
weights <- dist_weights(earns = earns, rho = 1.3, summary = TRUE, standardise = FALSE)
```
```{r, fig.align="center", echo = FALSE}
knitr::kable(weights) %>% 
  kable_styling(position = "center")
```

Because the distribution of earnings is approximately log-normal and so right skewed, the median is _less than_ the mean. As a result re-weighting using the welfare weights would increase the total utility (in this example the total utility gain is £`r format(round(sum(dist_weights(earns = earns, rho = 1.3, summary = FALSE, standardise = FALSE)$weight),0), scientific=FALSE)` for a £1 increase in income for all 10,000 individuals)

We can standardise the weights such that the mean weight is 1 and the total utility gain will equal the total income gain. In other words, the weighting does not effect the total utility.

```{r, echo = TRUE}
std_weights <- dist_weights(earns = earns, rho = 1.3, summary = TRUE, standardise = TRUE)
```
```{r, fig.align="center", echo = FALSE}
knitr::kable(std_weights) %>% 
  kable_styling(position = "center")
```


```{r, echo = FALSE}
# df <- dist_weights(earns = rnorm(n = 10000, 30000, 8000), rho = 1, summary = FALSE, standardise = TRUE)
# 
# summary(df$weight)
# sum(df$weight)
# 
# summary(df$std_weight)
# sum(df$std_weight)
```

# Flags / Potential Issues

- @layard_marginal_2008 use household income in their estimation, however, we have estimated individual level income
- Welfare weights are typically applied to just the bottom quintile not the whole distribution

# References

